Memory for television episodes preserves event content while introducing new across-event similarities

### 2019-12-11 paper/main.tex
updated topic modeling methods

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/topics}
\caption{\small \textbf{Transforming experience into memory.} \textbf{A.} Average correlations (across participants) between the topic vectors from each video event and the closest-matching recall events.  Error bars denote bootstrap-derived across-participant 95\% confidence intervals.  The stars denote the three best-remembered events (green) and worst-remembered events (red).  \textbf{B.} Wordles comprising the top 200 highest-weighted words reflected in the weighted-average topic vector across video events.  Green: video events were weighted by how well the topic vectors derived from recalls of those events matched the video events' topic vectors (Panel A).  Red: video events were weighted by the inverse of how well their topic vectors matched the recalled topic vectors.  \textbf{C.}  The set of all video and recall events is projected onto the two-dimensional space derived in Figure~\ref{fig:trajectory}.  The dots outlined in black denote video events (dot size reflects the average correlation between the video event's topic vector and the topic vectors from the closest matching recalled events from each participant; bigger dots denote stronger correlations).  The dots without black outlines denote recalled events.  All dots are colored using the same scheme as Figure~\ref{fig:trajectory}A.  Wordles for several example events are displayed (green: three best-remembered events; red: three worst-remembered events).  Within each circular wordle, the left side displays words associated with the topic vector for the video event, and the right side displays words associated with the (average) recall event topic vector, across all recall events matched to the given video event.}
\label{fig:topics}
\end{figure}

@@

% NOTE: for the "nearby moments of real life/naturalistic stimuli tend to be semantically similar" argument, is it worth adding something to the effect of:
% 1. we observed high temporal clustering scores
% 2. we also observed high semantic clustering scores
% 3. ergo, for both to be true, temporally proximal video events must be semantically similar
% - Paxton

Thus, analyses such as those in Figure \ref{fig:list-learning} that address only the temporal dynamics of free recall paint an incomplete picture of memory for naturalistic episodes.  While useful for studying presentation order-dependent recall dynamics, they neglect to consider the stimuli's content (or, for example, that content's potential interrelatedness).  However, sensitivity to stimulus and recall content introduces a new challenge: distinguishing between levels of recall quality for a stimulus (i.e., an event) that is considered to have been ``remembered."  When modeling memory experiments, often times events (or items) and their later memories are treated as binary and independent events (e.g., a given list item was simply either remembered or not remembered).  Various models of memory (e.g., \citealp{Yone02}) attempt to improve upon this by including confidence ratings, rendering this binary judgement instead categorical.  Our novel framework allows one to assess memory performance in a more continuous way (\textit{precision}), as well as analyze the correlational structure of each encoding event to each memory event (\textit{distinctiveness}). Further and importantly, these two novel metrics we introduce here arise from comparisons of the actual content of the experience/memories, which is not typically modeled.  Leveraging this, we find that the successful memory performance is related to 1) the precision with which the participant recounts each event and 2) the distinctiveness of each recall event (relative to the other recalled events). The first finding suggests that the information retained for \textit{any individual event} may predict the overall amount of information retained by the participant.  The second finding suggests that the ability to distinguish between temporally or semantically similar content is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore whether and how these metrics compare between cognitively impoverished groups and healthy controls.

@@

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised hand-generated annotations of each of 1000 scenes spanning the video clip~\citep[generated by][]{ChenEtal17}.  The features annotated included: narrative details (a sentence or two describing what happened in that scene); whether the scene took place indoors or outdoors; names of any characters that appeared in the scene; name(s) of characters in camera focus; name(s) of characters who were speaking in the scene; the location (in the story) that the scene took place; camera angle (close up, medium, long, top, tracking, over the shoulder, etc.); whether music was playing in the scene or not; and a transcription of any on-screen text.  We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene.  We then re-organized the text descriptions into overlapping sliding windows spanning 50 scenes each. In other words, the first text sample comprised the combined text from the first 50 scenes (i.e., 1--50), the second comprised the text from scenes 2--51, and so on.  We trained our model using these overlapping text samples with \texttt{scikit-learn}~\citep[version 0.19.1; ][]{PedrEtal11}, called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  Specifically, we used the \texttt{CountVectorizer} class to transform the text from each scene into a vector of word counts (using the union of all words across all scenes as the ``vocabulary,'' excluding English stop words); this yielded a number-of-scenes by number-of-words \textit{word count} matrix.  We then used the \texttt{LatentDirichletAllocation} class (topics=100, method=`batch') to fit a topic model~\citep{BleiEtal03} to the word count matrix, yielding a number-of-scenes (1000) by number-of-topics (100) \textit{topic proportions} matrix.  The topic proportions matrix describes which mix of topics (latent themes) is present in each scene.  Next, we transformed the topic proportions matrix to match the 1976 fMRI volume acquisition times.  For each fMRI volume, we took the topic proportions from whatever scene was displayed for most of that volume's 1500~ms acquisition time.  This yielded a new number-of-TRs (1976) by number-of-topics (100) topic proportions matrix.

We created similar topic proportions matrices using hand-annotated transcripts of each participant's recall of the video~\citep[annotated by ][]{ChenEtal17}.  We tokenized the transcript into a list of sentences, and then re-organized the list into overlapping sliding windows spanning 10 sentences each; in turn we transformed each window's sentences into a word count vector (using the same vocabulary as for the video model).  We then used the topic model already trained on the video scenes to compute the most probable topic proportions for each sliding window.  This yielded a number-of-sentences (range: 68--294) by number-of-topics (100) topic proportions matrix, for each participant.  These reflected the dynamic content of each participant's recalls.  Finally, we resampled each recall model to match the timecourse of the video model.  Note: for details on how we selected the video and recall window lengths and number of topics, see \textit{Supporting Information} and Figure~\topicopt.

@@

%\bibliography{memlab}
\bibliography{CDL-bibliography/memlabr}

>>>>

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/topics}
\caption{\small \textbf{Transforming experience into memory.} \textbf{A.} Average correlations (across participants) between the topic vectors from each video event and the closest-matching recall events.  Error bars denote bootstrap-derived across-participant 95\% confidence intervals.  The stars denote the three best-remembered events (green) and worst-remembered events (red).  \textbf{B.} Wordles comprising the top 200 highest-weighted words reflected in the weighted-average topic vector across video events.  Green: video events were weighted by how well the topic vectors derived from recalls of those events matched the video events' topic vectors (Panel A).  Red: video events were weighted by the inverse of how well their topic vectors matched the recalled topic vectors.  \textbf{C.}  The set of all video and recall events is projected onto the two-dimensional space derived in Figure~\ref{fig:trajectory}.  The dots outlined in black denote video events (dot size reflects the average correlation between the video event's topic vector and the topic vectors from the closest matching recalled events from each participant; bigger dots denote stronger correlations).  The dots without black outlines denote recalled events.  All dots are colored using the same scheme as Figure~\ref{fig:trajectory}A.  Wordles for several example events are displayed (green: three best-remembered events; red: three worst-remembered events).  Within each circular wordle, the left side displays words associated with the topic vector for the video event, and the right side displays words associated with the (average) recall event topic vector, across all recall events matched to the given video event.}
\label{fig:topics}
\end{figure}
% ------ NOTE: ------
% will also update this fig when new version is finished
% - Paxton

@@

Thus, analyses such as those in Figure \ref{fig:list-learning} that address only the temporal dynamics of free recall paint an incomplete picture of memory for naturalistic episodes.  While useful for studying presentation order-dependent recall dynamics, they neglect to consider the stimuli's content (or, for example, that content's potential interrelatedness).  However, sensitivity to stimulus and recall content introduces a new challenge: distinguishing between levels of recall quality for a stimulus (i.e., an event) that is considered to have been ``remembered."  When modeling memory experiments, often times events (or items) and their later memories are treated as binary and independent events (e.g., a given list item was simply either remembered or not remembered).  Various models of memory (e.g., \citealp{Yone02}) attempt to improve upon this by including confidence ratings, rendering this binary judgement instead categorical.  Our novel framework allows one to assess memory performance in a more continuous way (\textit{precision}), as well as analyze the correlational structure of each encoding event to each memory event (\textit{distinctiveness}). Further and importantly, these two novel metrics we introduce here arise from comparisons of the actual content of the experience/memories, which is not typically modeled.  Leveraging this, we find that the successful memory performance is related to 1) the precision with which the participant recounts each event and 2) the distinctiveness of each recall event (relative to the other recalled events). The first finding suggests that the information retained for \textit{any individual event} may predict the overall amount of information retained by the participant.  The second finding suggests that the ability to distinguish between temporally or semantically similar content is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore whether and how these metrics compare between cognitively impoverished groups and healthy controls.
% ------ NOTE: ------
% need to update text here to reflect new distinctivness stats in precision/distinctiveness fig
% - Paxton

@@

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised 998 hand-generated annotations of short (mean: 2.96s) scenes spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two referring to the break between the first and second scan sessions, during which no fMRI data was collected).  The features annotated included: narrative details (a sentence or two describing what happened in that scene); whether the scene took place indoors or outdoors; names of any characters that appeared in the scene; name(s) of characters in camera focus; name(s) of characters who were speaking in the scene; the location (in the story) that the scene took place; camera angle (close up, medium, long, top, tracking, over the shoulder, etc.); whether music was playing in the scene or not; and a transcription of any on-screen text.  We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene.  We then re-organized the text descriptions into overlapping sliding windows of 50 scenes each.  In other words, we created a ``context" for each scene comprising the text descriptions of the preceding 25 scenes, the present scene, and the following 24 scenes.  To model the ``context" at the beginning and end of the video (i.e., within 50 scenes of the beginning or end), we created overlapping sliding windows that grew in size from one scene to the full length, then similarly tapered their length at the end.  This bore the additional benefit of representing each scene's description in the text corpus an equal number of times.

We trained our model using these overlapping text samples with \texttt{scikit-learn}~\citep[version 0.19.1; ][]{PedrEtal11}, called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  Specifically, we used the \texttt{CountVectorizer} class to transform the text from each window into a vector of word counts (using the union of all words across all scenes as the ``vocabulary,'' excluding English stop words); this yielded a number-of-windows by number-of-words \textit{word count} matrix.  We then used the \texttt{LatentDirichletAllocation} class (topics=100, method=`batch') to fit a topic model~\citep{BleiEtal03} to the word count matrix, yielding a number-of-windows (1047) by number-of-topics (100) \textit{topic proportions} matrix.  The topic proportions matrix describes which mix of topics (latent themes) is present in and around each scene.  Next, we transformed the topic proportions matrix to match the 1976 fMRI volume acquisition times.  We assigned each topic vector to the timepoint midway between the beginning of the first scene and the end of the last scene in its corresponding sliding text window.  We then transformed these timepoints to units of TRs and interpolated the dynamic topic proportions matrix to obtain number-of-TRs (1976) by number-of-topics (100) matrix.

We created similar topic proportions matrices using hand-annotated transcripts of each participant's recall of the video~\citep[annotated by ][]{ChenEtal17}.  We tokenized the transcript into a list of sentences, and then re-organized the list into overlapping sliding windows spanning 10 sentences each (and analogously tapered the lengths of the first and last 10 sliding windows).  In turn, we transformed each window's sentences into a word count vector (using the same vocabulary as for the video model).  We then used the topic model already trained on the video scenes to compute the most probable topic proportions for each sliding window.  This yielded a number-of-windows (range: 83--312) by number-of-topics (100) topic proportions matrix for each participant.  These reflected the dynamic content of each participant's recalls.  Note: for details on how we selected the video and recall window lengths and number of topics, see \textit{Supporting Information} and Figure~\topicopt.

@@

\bibliography{../../CDL-bibliography/memlab}
%\bibliography{CDL-bibliography/memlab}
